A weakly structured stem for human origins in Africa

### 2023-03-03 paper.tex
Update paper.tex

Evidence for archaic hominin admixture in Eurasia has bolstered the plausibility of 
archaic hominin admixture having also occurred in Africa. Previous work that sought to 
explain patterns of polymorphism inconsistent with a single-origin model thus focused on 
archaic hominin admixture as an alternative model, by referring to additional 
(or ``ghost'') branches required to fit the data as ``archaic'' 
\citep{Plagnol2006-lt,Hammer2011-bx,Hey2018-pw,Ragsdale2019-nt,Lorente-Galdos2019-vz,Durvasula2020-td} 
and assuming (or inferring) deep divergences. This perspective has oriented interpretations 
of both genomic (e.g., selection \cite{Wall2019-ao}) and fossil (e.g., evolution of 
\emph{Homo sapiens} \cite{Grun2020-su} and late survival of archaic hominins and gene 
flow from them ADD Harvati et al. 2011 ``The Later Stone Age Calvaria from 
Iwo Eleru, Nigeria: Morphology and Chronology'', Crevecoeur 2012 ``The Upper Paleolithic 
Human Remains of Nazlet Khater 2 (Egypt) and Past Modern Human Diversity'', Crevecoeur 
et al. 2016 ``Late Stone Age Human Remains from Ishango (Democratic Republic of Congo): 
New Insights on Late Pleistocene Modern Human Diversity in Africa''}). Here, we have shown 
that a model of a weakly structured stem better captures the apparently inconsistent 
patterns of polymorphisms.

Preferring weakly-structured-stem models over archaic-admixture models potentially has 
a range of implications. First, with a weakly structured stem, there is no need to posit 
that an archaic hominin population in Africa stayed reproductively isolated from the 
ancestral human lineage for hundreds of thousands of years before the initiation of 
gene flow. Instead, there would simply have been continuous or recurrent contact between 
two or more groups present in Africa. Additionally, because we would have expected to 
detect admixture from late surviving archaic hominins, the distinctive morphologies of 
certain fairly recent \emph{Homo sapiens} fossils (ADD Harvati et al. 2011 
``The Later Stone Age Calvaria from Iwo Eleru, Nigeria: Morphology and Chronology'', 
Crevecoeur 2012 ``The Upper Paleolithic Human Remains of Nazlet Khater 2 (Egypt) and 
Past Modern Human Diversity'', Crevecoeur et al. 2016 ``Late Stone Age Human Remains 
from Ishango (Democratic Republic of Congo): New Insights on Late Pleistocene Modern 
Human Diversity in Africa''}) may have causes other than admixture \citep{Bergstrom2021-iw.

>>>>

Evidence for archaic hominin admixture in Eurasia has bolstered the plausibility of 
archaic hominin admixture having also occurred in Africa. Previous work that sought to 
explain patterns of polymorphism inconsistent with a single-origin model thus focused on 
archaic hominin admixture as an alternative model, by referring to additional 
(or ``ghost'') branches required to fit the data as ``archaic'' 
\citep{Plagnol2006-lt,Hammer2011-bx,Hey2018-pw,Ragsdale2019-nt,Lorente-Galdos2019-vz,Durvasula2020-td} 
and assuming (or inferring) deep divergences. This perspective has oriented interpretations 
of both genomic (e.g., selection \cite{Wall2019-ao}) and fossil (e.g., evolution of 
\emph{Homo sapiens} \cite{Grun2020-su} and late survival of archaic hominins and gene 
flow from them ADD Harvati et al. 2011 ``The Later Stone Age Calvaria from 
Iwo Eleru, Nigeria: Morphology and Chronology'', Crevecoeur 2012 ``The Upper Paleolithic 
Human Remains of Nazlet Khater 2 (Egypt) and Past Modern Human Diversity'', Crevecoeur 
et al. 2016 ``Late Stone Age Human Remains from Ishango (Democratic Republic of Congo): 
New Insights on Late Pleistocene Modern Human Diversity in Africa''). Here, we have shown 
that a model of a weakly structured stem better captures the apparently inconsistent 
patterns of polymorphisms.

Preferring weakly-structured-stem models over archaic-admixture models potentially has 
a range of implications. First, with a weakly structured stem, there is no need to posit 
that an archaic hominin population in Africa stayed reproductively isolated from the 
ancestral human lineage for hundreds of thousands of years before the initiation of 
gene flow. Instead, there would simply have been continuous or recurrent contact between 
two or more groups present in Africa. Additionally, because we would have expected to 
detect admixture from late surviving archaic hominins, the distinctive morphologies of 
certain fairly recent \emph{Homo sapiens} fossils (ADD Harvati et al. 2011 
``The Later Stone Age Calvaria from Iwo Eleru, Nigeria: Morphology and Chronology'', 
Crevecoeur 2012 ``The Upper Paleolithic Human Remains of Nazlet Khater 2 (Egypt) and 
Past Modern Human Diversity'', Crevecoeur et al. 2016 ``Late Stone Age Human Remains 
from Ishango (Democratic Republic of Congo): New Insights on Late Pleistocene Modern 
Human Diversity in Africa'') may have causes other than admixture \citep{Bergstrom2021-iw}.
